Kinase similarity assessment pipeline for off-target prediction [Article v\versionnumber]

### 2022-06-22 manuscript/manuscript.tex
Fix typos/grammar/etc..

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR TRAINING ARTICLE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,training]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% Use the "ASAPversion' option following article acceptance to add the DOI and relevant dates to the document footer.
% Use the 'pubversion' option for adding the citation and publication information to the document footer, when the LiveCoMS issue is finalized.
% The 'training' option for indicates that this is a training article.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

@@

\author[1\authfn{1}]{Talia B. Kimber}
\author[1\authfn{1}\authfn{2}]{Dominique Sydow}
\author[1*]{Andrea Volkamer}
\affil[1]{\textit{In silico} Toxicology and Structural Bioinformatics, Institute of Physiology, Charit\'e-Universit\"atsmedizin Berlin, Charit\'eplatz 1, 10117, Berlin, Germany}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubissue{<issue>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Day Month Year}
\dateaccepted{Day Month Year}

@@

\begin{abstract}
Kinases are established drug targets to combat cancer and inflammatory diseases. Despite decades of kinase research, challenges still remain, such as the under-exploration of a large fraction of the kinome and the promiscuous binding of many kinase inhibitors. Due to the highly conserved orthosteric ATP binding site in kinases, ligands may bind not only to their designated kinase (on-target) but also to other kinases (off-targets). Such promiscuous binding can cause mild to severe side effects, and the prediction of these off-targets is highly non-trivial.
Therefore, we propose a pipeline that allows the study of kinase similarities from four different angles in an automated and modular fashion. The first method considers the binding site sequence. The second method uses structural information via KiSSim, a newly developed fingerprint that considers both physico-chemical and spatial properties of the binding site. The third method involves kinase-ligand interaction fingerprints as provided by KLIFS, and the last method utilizes the measured activity of ligands on kinases based on ChEMBL data. Finally, results for a given set of kinases are collected and analyzed to gain insight into potential off-targets from the different aforementioned perspectives. Since the pipeline is set up as a series of Jupyter notebooks covering both theoretical and practical aspects, the target audience ranges from beginners to advanced users working in the field of natural and computer sciences. The pipeline is part of the TeachOpenCADD project and extends it with this special kinase edition. All code is free, open-source, and made available at \href{https://github.com/volkamerlab/teachopencadd
}{https://github.com/volkamerlab/teachopencadd
}.
\end{abstract}

@@

\section{Introduction}
Kinases are involved in most cellular processes by phosphorylating\textemdash and thereby activating\textemdash themselves or other proteins. This family is among the most frequently mutated proteins in tumors and has been successfully studied as drug targets for many decades~\cite{Cohen_2021_NatRevDrugDiscov}. Thanks to the longstanding research, a plethora of kinase data is freely available, i.e. as part of databases such as UniProt~\cite{uniprot_consortium_2020_nar}, PDB~\cite{Berman_2000_NAR} or ChEMBL~\cite{Gaulton_2016_nar}, and has been made easily accessible via kinase resources such as the KLIFS\textemdash Kinase-Ligand Interaction Fingerprints and Structures\textemdash database~\cite{Kanev_2020_NAR}. As of February 2022, $5\,911$ X-ray structures of human kinases have been resolved (see the KLIFS database~\cite{klifs_feb_2022}) and $70$ FDA-approved small molecule protein kinase inhibitors are on the market~\cite{FDA_PKI_2022}. Most of the approved drugs bind in the ATP-binding pocket and intermediate surroundings (orthosteric binding site).

Although structural information provides rich information, kinases have been widely classified based on sequence. \citet{Manning_2022_science} clustered the human protein kinases based on their sequence similarity into eight major groups (AGC, CAMK, CK1, CMGC, RGC, STE, TK, TKL) and one “Other” group for unassigned kinases, as well as atypical kinases. The resulting Manning kinome tree depicts kinase clustering (see Figure~\ref{fig:kinmap}).

Despite decades of kinase research, challenges still remain~\cite{Kooistra_2017_AnnualReportsMedicinalChemistry}. For example: \begin{enumerate}
    \item A large fraction of the kinome is un-/underexplored. Figure~\ref{fig:structure_per_kinase} shows the number of PDB structures per kinase, unveiling a vast imbalance between structurally resolved kinases and unexplored ones. For example, CDK2 has been resolved in $426$ PDB structures, while only $313$ kinases~\cite{klifs_feb_2022} out of approximately $540$ in the kinome~\cite{Kooistra_2017_AnnualReportsMedicinalChemistry} have been structurally resolved.
    \item Many kinase inhibitors are promiscuous binders causing off-target effects or enabling polypharmacology ~\cite{Cohen_2021_NatRevDrugDiscov, Morphy_2009_JMedChem}. For example, the Epidermal Growth Factor Receptor~(EGFR) inhibitor erlotinib shows affinities to other kinases in the highly sequentially-similar TK kinase group, but also strongly affects off-targets in more remote kinase groups (see Figure\ref{fig:off_target}).
\end{enumerate}

Therefore, assessing kinase similarity from different angles may be a crucial step in understanding and predicting off-targets to help designing more selective drugs and avoiding side effects.

\subsection{Scope}
In this study, similarities between a set of kinases are investigated based on methods offering different perspectives on this challenging topic with a focus on orthosteric binding sites (here referred to as binding sites), as summarized in Table~\ref{tab:notebook_overview}. The first method considers the binding site sequence as deposited in the KLIFS database. The second method uses KiSSim~\cite{sydow_2021_kissim_chemrxiv}, a recently developed fingerprint that considers physico-chemical as well as spatial properties of the binding site. The third method involves protein-ligand interaction fingerprints as provided in the KLIFS database, and the last method utilizes the measured activity of ligands against kinases based on ChEMBL data~\cite{Gaulton_2016_nar}. The different methods are preceded by a general introduction to kinases and the challenges faced during kinase-centric drug design, and succeeded by a comparison between the different kinase similarity methods.

Please note that this study focuses on the similarities between ATP binding sites. Therefore, kinase polypharmacology and off-targets can only be assessed within the scope of orthosteric binding sites, even though the promiscuity of some ligands may be explained by binding to allosteric binding sites (potential allosteric binding sites are summarized in the Kinase Atlas~\cite{Yueh_2019_JMedChem}).

This study has been put together into a modular pipeline that enables the research of kinase similarities in an automated fashion, allowing users to simply use it out of the box, or adapt it to their needs.

@@

\begin{figure*}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{kinmap_n_structures_per_kinase.png}
         \caption{Number of PDB structures per kinase. The figure shows the imbalance between highly explored kinases, e.g. groups: TK, CMGC, and CAMK, and less explored ones, e.g. the CK1 group. CDK2 has the most structures, with $426$. The red circle is proportional to the number of PDB structures for each kinase, such that the greater is the circle, the higher is the number of structures.}
         \label{fig:structure_per_kinase}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{kinmap_erlotinib_karaman.png}
         \caption{Developing selective kinase inhibitors is non-trivial since kinases are highly conserved in the ATP binding site. EGFR inhibitor erlotinib binds not only to its intended target EGFR, but also to kinases in remote groups, such as SLK/LOK in the STE group and GAK in the CMGC group. The blue circle is proportional to the $K_d$ value in \SI{}{\nano\Molar} taken from the \citet{Karaman_2008_NatBiotechnol} dataset.}
         \label{fig:off_target}
     \end{subfigure}
        \caption{Visual representation using the Manning tree of existing challenges in kinase research: un-/underexplored kinase groups (left) and the promiscuity character of kinases (right). The figure is taken from \href{https://projects.volkamerlab.org/teachopencadd/talktorials/T023\_what\_is\_a\_kinase.html}{https://projects.volkamerlab.org/teachopencadd/talktorials/T023\_what\_is\_a\_kinase.html} and is generated using KinMap~\cite{Eid_2017_BMCBioinformatics}.}
        \label{fig:kinmap}
\end{figure*}

This workflow is integrated in the context of TeachOpenCADD \cite{Sydow_2019_JCheminform, sydow_2022_nar}, a teaching platform for computer-aided drug design (CADD) using open-source packages and data. Specific tasks in cheminformatics and structural bioinformatic are described and solved using Python-based Jupyter notebooks \cite{Kluyver_2016_Jupyter} as interacting platform. All code has been deposited on GitHub, see
\\
\href{https://github.com/volkamerlab/teachopencadd}{https://github.com/volkamerlab/teachopencadd}. And the project website can be found at this link,
\\
\href{https://projects.volkamerlab.org/teachopencadd/}{https://projects.volkamerlab.org/teachopencadd/}. 
\section{Prerequisites}
\subsection{Target audience}
The notebooks were developed to support researchers interested in kinase-centric computational drug design with a focus on understanding and predicting kinase off-targets. 
As this collection is part of the TeachOpenCADD~\cite{Sydow_2019_JCheminform, sydow_2022_nar} training material, we also recommend the notebooks to teachers as pedagogical interactive material in structural bioinformatics and cheminformatics.

@@

\subsection{Software requirements}
The notebooks are written in Python and rely on open-source packages such as pandas~\cite{pandas_2020}, numpy~\cite{harris_2020_numpy}, scipy~\cite{Virtanen_2020_NMeth}, matplotlib~\cite{Hunter_2007_IEEE}, seaborn~\cite{Waskom_2021_seaborn}, scikit-learn~\cite{Pedregosa_2011_JMLR}, rdkit~\cite{RDKit_2022}, biotite~\cite{Kunzmann_2018_biotite}, opencadd~\cite{Sydow_2022_JOSS}, requests~\cite{requests_2022}, and pillow~\cite{pillow_2022}.

@@

\section{Method}
\label{sec:method}
In this section, the four methods that are introduced to quantify kinase similarity are described, namely the pocket sequence, the KiSSim fingerprint, the interaction fingerprint, and the ligand profile. Note that the theoretical and practical aspects of each method are also covered in great detail in the individual notebooks of this kinase collection (Table~\ref{tab:notebook_overview}). As discussed in the "Scope" section of this manuscript, we focus on kinase similarities based on orthosteric binding sites.

@@

\subsubsection{Sequence similarity}
Unlike sequence identity which treats all residues uniformly, pairwise sequence similarity, or sequence similarity, takes into account the change of the amino acids over evolutionary time, thus reflecting relationships between amino acids. It is based on a substitution matrix $M$, where each entry gives a score between two amino acids. In this study, the BLOSUM substitution matrix~\cite{Henikoff_1992_PNAS}, as implemented in biotite~\cite{Kunzmann_2018_BMC}, is used. Formally, the following is defined:

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\columnwidth]{sequence_similarity.png}
    \caption{Sequence similarity between EGFR and MET. The $85$ residue pocket sequence is retrieved from KLIFS. The pairwise sequence similarity takes into account the change of the amino acids over evolutionary time.}
    \label{fig:pocket_sequence}
\end{figure}

\subsection{The KiSSim fingerprint}
In order to assess the pairwise similarity of kinases from a structural point of view, the newly developed KiSSim (\textbf{Ki}nase \textbf{S}tructure \textbf{Sim}ilarity) fingerprint~\cite{sydow_2021_kissim_chemrxiv, kissim_package} is used. This fingerprint describes the physico-chemical and spatial properties of structurally resolved kinases, while focusing on the KLIFS pocket residues. Each structure is mapped to a fingerprint composed of $1032$ bits, the first $680 \, (=85 \times 8)$ describing physico-chemical features and the remaining $352 \, (=85 \times 4 + 12)$ spatial information (see Figure~\ref{fig:kissim_similarity}).

\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{kissim_fingerprint.png}
    \caption{The $1032$-long KiSSim fingerprint encodes physico-chemical and spatial properties of the kinase's pocket, adding a structural perspective on kinases. The figure is adapted from \cite{kissim_package}.}
    \label{fig:kissim_similarity}
\end{figure}

@@

\subsection{The interaction fingerprint}
Interaction fingerprints~(IFPs) encode the binding mode of a ligand in a binding site, i.e., the protein-ligand interactions that are present in a structurally resolved complex. If a ligand can form similar interaction patterns in proteins other than its designated protein (off- vs. on-target), it is possible that this ligand will cause unintended side effects. Knowledge about binding mode similarities can therefore help to avoid such off-target effects.
% ; or to exploit the known similarities for polypharmacology effects where one ligand could intentionally target multiple proteins at once.

@@

\subsubsection{Bioactivity data}
The bioactivity data used for this method comes from Kinodata~\cite{kinodata_2022}, from the Openkinome organization~\cite{openkinome_feb_2022}. It is a pre-processed kinase subset of the ChEMBL data~\cite{Gaulton_2016_nar}, version 29. Further processing includes keeping only $IC_{50}$ values given in \SI{}{\nano\Molar}, and converted to $pIC_{50}$s. If there are several measurements for a kinase-compound pair, then the most active value, i.e. the entry with the highest $pIC_{50}$, value is kept. Finally, the $pIC_{50}$ values are binarized using a $6.3$ cutoff to discriminate between an active or inactive compound as described in~\cite{Merget_2017_JMedChem}.

In the pipeline, one can additionally compute the non-reduced ratio of number of active compound against the total number of compounds to gain insight into the actual number of measurements for each kinase pair.

@@

\subsubsection{Heatmaps}
The heatmaps are generated using Matplotlib~\cite{Hunter_2007_IEEE} to depict the similarity between a set of kinases. The maximum value is $1$, indicating exact similarity, as is the case for diagonal entries. The value $0$ indicates total dissimilarity. Plotting such figures allows to see and extract patterns thanks to the gradient of colors, see top row in Figure~\ref{fig:comparison}.

\subsubsection{Dendrograms}
Clustering algorithms are used to identify groups such that the similarities within clusters are higher than compared to other clusters~\cite{Hastie_2009_ESLii}. In this study, hierarchical clustering is used, and, unlike heatmaps, it is based on distance (or dissimilarity).
% According to \citet{Hastie_2009_ESLii}, "The goal of cluster analysis is to partition the observations into groups (“clusters”) so that the pairwise dissimilarities between those assigned to the same cluster tend to be smaller than those in different clusters." 
Hierarchical clustering can be graphically displayed using a dendrogram (see bottom row in Figure~\ref{fig:comparison}), where the height of each node is proportional to the dissimilarity between its two daughter clusters.
% the height of each node is proportional to the value of the intergroup dissimilarity between its two daughters.
The clustering and plotting is done using Scikit-learn~\cite{Pedregosa_2011_JMLR} and Matplotlib~\cite{Hunter_2007_IEEE}, respectively.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{all_comparison.png}
    \caption{Visualization of kinase similarity from four different angles: sequence, KiSSim, interaction fingerprint~(ifp) as well as ligand profile. The top, bottom row shows four heatmaps, dendrograms respectively for a set of nine study kinases.}
    \label{fig:comparison}
\end{figure*}

@@

In the pipeline presented herein, we implement the different methods once and streamline each method's results into a standardized output with a pre-defined set of visualization tools for easy comparison, while leaving the pipeline flexible enough so that adding new methods or new visualization tools is effortless, making the whole process easy to understand, maintain, and expand.

\subsection{Means of the pipeline}
The proposed pipeline is a collection of six Jupyter notebooks~\cite{Kluyver_2016_Jupyter} that allows the study of kinase similarities from four different angles in an automated and modular fashion (Figure~\ref{fig:pipeline}).

@@

\subsection{Structure of the notebooks}
The structure of all notebooks is as follows: the first section covers the theory written in Markdown and summarizes the necessary concepts to understand the task. Relevant references are also mentioned. The second part of a notebook deals with the actual implementation of the task in a pedagogical manner, including motivation for practical steps and detailed comments on coding decisions. Finally, a discussion and a quiz section wrap up the notebook.
This structure is very well suited from a teaching perspective, since it contains both theory and hands on programming. The notebook can easily be used as a medium for a presentation, and it allows for self-study and usage in own research projects.

\subsection{About the code}
% Python and pythonic code
The programming part is done in Python exclusively and the code follows the latest software best practices. It is written pythonically and contains lots of code comments.
% reproducibility
Thanks to the continuous integration~(CI), all outputs and results are fully reproducible and the pipeline's maintenance is facilitated.

@@

The final notebook collects the information from the previous ones and compares the different perspectives with easy-to-understand visualization such as heatmaps and dendrograms (see Section~\ref{sec:comparison}). Additionally, an equally weighted average to combine distance and similarity matrices of all four perspectives can be computed, yielding a single heatmap, and a single dendrogram. The user can easily extend this to a knowledge-informed weighted scheme based on their own research focus.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR TRAINING ARTICLE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,training,ASAPversion]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% Use the "ASAPversion' option following article acceptance to add the DOI and relevant dates to the document footer.
% Use the 'pubversion' option for adding the citation and publication information to the document footer, when the LiveCoMS issue is finalized.
% The 'training' option for indicates that this is a training article.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

@@

\author[1,2\authfn{1}]{Talia B. Kimber}
\author[1\authfn{1}\authfn{2}]{Dominique Sydow}
\author[1*]{Andrea Volkamer}
\affil[1]{\textit{In silico} Toxicology and Structural Bioinformatics, Institute of Physiology, Charit\'e-Universit\"atsmedizin Berlin, Charit\'eplatz 1, 10117, Berlin, Germany}
\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, New York 10065, United States}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubissue{<issue>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{17 March 2022}
\dateaccepted{15 June 2022}

@@

\begin{abstract}
Kinases are established drug targets to combat cancer and inflammatory diseases. Despite decades of kinase research, challenges still remain, such as the under-exploration of a large fraction of the kinome and the promiscuous binding of many kinase inhibitors. Due to the highly conserved orthosteric ATP binding site in kinases, ligands may bind not only to their designated kinase (on-target) but also to other kinases (off-targets). Such promiscuous binding can cause mild to severe side effects, and the prediction of these off-targets is highly non-trivial.
Therefore, we propose a pipeline that allows the study of kinase similarities from four different angles in an automated and modular fashion. The first method considers the binding site sequence. The second method uses structural information via KiSSim, a newly developed fingerprint that considers both physico-chemical and spatial properties of the binding site. The third method involves kinase-ligand interaction fingerprints as provided by KLIFS, and the last method utilizes the measured activity of ligands on kinases based on ChEMBL data. Finally, results for a given set of kinases are collected and analyzed to gain insight into potential off-targets from the different aforementioned perspectives. Since the pipeline is set up as a series of Jupyter notebooks covering both theoretical and practical aspects, the target audience ranges from beginners to advanced users working in the field of natural and computer sciences. The pipeline is part of the TeachOpenCADD project and extends it with this special kinase edition. All code is free, open-source, and made available at \href{https://github.com/volkamerlab/teachopencadd}{https://github.com/volkamerlab/teachopencadd}.
\end{abstract}

@@

\section{Introduction}
Kinases are involved in most cellular processes by phosphorylating\textemdash and thereby activating\textemdash themselves or other proteins. This family is among the most frequently mutated proteins in tumors and kinases have been successfully studied as drug targets for many decades~\cite{Cohen_2021_NatRevDrugDiscov}. Thanks to the longstanding research, a plethora of kinase data is freely available, i.e., as part of databases such as UniProt~\cite{uniprot_consortium_2020_nar}, PDB~\cite{Berman_2000_NAR} or ChEMBL~\cite{Gaulton_2016_nar}, and has been made easily accessible via kinase resources such as the KLIFS\textemdash Kinase-Ligand Interaction Fingerprints and Structures\textemdash database~\cite{Kanev_2020_NAR}. As of February 2022, $5,911$ X-ray structures of human kinases have been resolved (see the KLIFS database~\cite{klifs_feb_2022}) and $70$ FDA-approved small molecule protein kinase inhibitors are on the market~\cite{FDA_PKI_2022}. Most of the approved drugs bind in the ATP binding pocket and intermediate surroundings (orthosteric binding site).

Although structural data provides rich information, kinases have been widely classified based on sequence. \citet{Manning_2022_science} clustered the human protein kinases based on their sequence similarity into eight major groups (AGC, CAMK, CK1, CMGC, STE, TK, TKL, and "Other") as well as atypical kinases. The resulting Manning kinome tree depicts kinase clustering (see Figure~\ref{fig:kinmap}).

Despite decades of kinase research, challenges still remain~\cite{Kooistra_2017_AnnualReportsMedicinalChemistry}. For example: \begin{enumerate}
    \item A large fraction of the kinome is un-/underexplored. Figure~\ref{fig:structure_per_kinase} shows the number of PDB structures per kinase, unveiling a vast imbalance between structurally resolved kinases and unexplored ones. For example, CDK2 has been resolved in $426$ PDB structures, while only $313$ kinases~\cite{klifs_feb_2022} out of approximately $540$ in the kinome~\cite{Kooistra_2017_AnnualReportsMedicinalChemistry} have been structurally resolved.
    \item Many kinase inhibitors are promiscuous binders, causing off-target effects or enabling polypharmacology ~\cite{Cohen_2021_NatRevDrugDiscov, Morphy_2009_JMedChem}. For example, the Epidermal Growth Factor Receptor~(EGFR) inhibitor erlotinib shows affinities to other kinases in the highly sequentially-similar TK kinase group, but also strongly affects off-targets in more remote kinase groups (see Figure\ref{fig:off_target}).
\end{enumerate}

Therefore, assessing kinase similarity from different angles may be a crucial step in understanding and predicting off-targets to help to design more selective drugs and to avoid side effects.

\subsection{Scope}
In this study, similarities between a set of kinases are investigated based on methods offering different perspectives on this challenging topic with a focus on orthosteric binding sites (here referred to as binding sites), as summarized in Table~\ref{tab:notebook_overview}. The first method considers the binding site sequence as deposited in the KLIFS database. The second method uses KiSSim~\cite{sydow_2022_jcim}, a recently developed fingerprint that considers physico-chemical as well as spatial properties of the binding site. The third method involves protein-ligand interaction fingerprints as provided in the KLIFS database, and the last method utilizes the measured activity of ligands against kinases based on ChEMBL data~\cite{Gaulton_2016_nar}. The different methods are preceded by a general introduction to kinases and the challenges faced in kinase-centric drug design, and succeeded by a comparison between the different kinase similarity methods.

Note that this study focuses on the similarities between ATP binding sites. Therefore, kinase polypharmacology and off-targets can only be assessed within the scope of orthosteric binding sites, even though the promiscuity of some ligands may be explained by binding to allosteric binding sites. Potential allosteric binding sites are summarized in the Kinase Atlas~\cite{Yueh_2019_JMedChem}.

This study has been assembled into a modular pipeline that enables the research of kinase similarities in an automated fashion, allowing users to simply use it out of the box, or adapt it to their needs.

@@

\begin{figure*}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{kinmap_n_structures_per_kinase.png}
         \caption{Number of PDB structures per kinase. The figure shows the imbalance between highly explored kinases, for example, the groups TK and CMGC. The CDK2 kinase in the CMGC group has the most structures, with $426$. The red circle is proportional to the number of PDB structures for each kinase, such that the greater is the circle, the higher is the number of structures.}
         \label{fig:structure_per_kinase}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{kinmap_erlotinib_karaman.png}
         \caption{Developing selective kinase inhibitors is non-trivial since kinases are highly conserved in the ATP binding site. EGFR inhibitor erlotinib binds not only to its intended target EGFR, but also to kinases in remote groups, such as SLK/LOK in the STE group and GAK in the "Other" group. The blue circle is proportional to the $K_d$ value in \SI{}{\nano\Molar} taken from the \citet{Karaman_2008_NatBiotechnol} dataset.}
         \label{fig:off_target}
     \end{subfigure}
        \caption{Visual representation using the Manning tree of existing challenges in kinase research: un-/underexplored kinase groups (left) and the promiscuous behavior of kinases (right). The figure is taken from \href{https://projects.volkamerlab.org/teachopencadd/talktorials/T023\_what\_is\_a\_kinase.html}{https://projects.volkamerlab.org/teachopencadd/talktorials/T023\_what\_is\_a\_kinase.html} and is generated using KinMap~\cite{Eid_2017_BMCBioinformatics}.}
        \label{fig:kinmap}
\end{figure*}

This workflow is integrated in the context of TeachOpenCADD \cite{Sydow_2019_JCheminform, sydow_2022_nar}, a teaching platform for computer-aided drug design (CADD) using open-source packages and data. Specific tasks in cheminformatics and structural bioinformatic are described and solved using Python-based Jupyter notebooks \cite{Kluyver_2016_Jupyter} as interactive platform. All code has been deposited on GitHub, see \url{https://github.com/volkamerlab/teachopencadd}. The project website can be found at this link, \url{https://projects.volkamerlab.org/teachopencadd}.

\section{Prerequisites}
\subsection{Target audience}
The notebooks were developed to support researchers interested in kinase-centric computational drug design, with a focus on understanding and predicting kinase off-targets. 
As this collection is part of the TeachOpenCADD training material~\cite{Sydow_2019_JCheminform, sydow_2022_nar}, we also recommend the notebooks to teachers as pedagogical interactive material in structural bioinformatics and cheminformatics.

@@

\subsection{Software requirements}
The notebooks are written in Python and rely on open-source packages such as pandas~\cite{pandas_2020}, numpy~\cite{harris_2020_numpy}, scipy~\cite{Virtanen_2020_NMeth}, matplotlib~\cite{Hunter_2007_IEEE}, seaborn~\cite{Waskom_2021_seaborn}, scikit-learn~\cite{Pedregosa_2011_JMLR}, rdkit~\cite{RDKit_2022}, biotite~\cite{Kunzmann_2018_biotite}, opencadd~\cite{Sydow_2022_JOSS}, kissim \cite{kissim_package}, and requests~\cite{requests_2022}.

@@

\section{Method}
\label{sec:method}
In this section, the four methods that are introduced to measure kinase similarity are described, namely the pocket sequence, the KiSSim fingerprint, the interaction fingerprint, and the ligand profile. Note that the theoretical and practical aspects of each method are also covered in great detail in the individual notebooks of this kinase collection (Table~\ref{tab:notebook_overview}). As discussed in the "Scope" section of this manuscript, we focus on kinase similarity based on orthosteric binding sites.

@@

\subsubsection{Sequence similarity}
Unlike sequence identity which treats all residues uniformly, pairwise sequence similarity, or sequence similarity, takes into account the change of the amino acids over evolutionary time, thus, reflecting relationships between amino acids. It is based on a substitution matrix $M$, where each entry gives a score between two amino acids. In this study, the BLOSUM substitution matrix~\cite{Henikoff_1992_PNAS}, as implemented in biotite~\cite{Kunzmann_2018_BMC}, is used. Formally, the following is defined:

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\columnwidth]{sequence_similarity.png}
    \caption{Sequence similarity between EGFR and MET. The $85$- residue pocket sequence is retrieved from KLIFS. The pairwise sequence similarity takes into account the change of the amino acids over evolutionary time.}
    \label{fig:pocket_sequence}
\end{figure}

\subsection{The KiSSim fingerprint}
In order to assess the pairwise similarity of kinases from a structural point of view, the newly developed KiSSim (\textbf{Ki}nase \textbf{S}tructure \textbf{Sim}ilarity) fingerprint~\cite{sydow_2022_jcim, kissim_package} is used. This fingerprint describes the physico-chemical and spatial properties of structurally resolved kinases, while focusing on the KLIFS pocket residues. Each structure is mapped to a fingerprint composed of $1,032$ bits, the first $680 \, (=85 \times 8)$ bits describing physico-chemical features and the remaining $352 \, (=85 \times 4 + 12)$ bits spatial information (see Figure~\ref{fig:kissim_similarity}).

\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{kissim_fingerprint.png}
    \caption{The $1,032$-long KiSSim fingerprint encodes physico-chemical and spatial properties of the kinase's pocket, adding a structural perspective on kinases. The figure is adapted from \cite{kissim_package}.}
    \label{fig:kissim_similarity}
\end{figure}

@@

\subsection{The interaction fingerprint}
Interaction fingerprints~(IFPs) encode the binding mode of a ligand in a binding site, i.e., the protein-ligand interactions that are present in a structurally resolved complex. If a ligand can form similar interaction patterns in proteins other than its designated protein (off- vs. on-target), it is possible that this ligand will cause unintended side effects. Knowledge about binding mode similarities can therefore help to avoid such off-target effects.

@@

\subsubsection{Bioactivity data}
The bioactivity data used for this method comes from Kinodata~\cite{kinodata_2022}, from the Openkinome organization~\cite{openkinome_feb_2022}. It is a pre-processed kinase subset of the ChEMBL data~\cite{Gaulton_2016_nar}, version 29. Further processing includes keeping only $IC_{50}$ values given in \SI{}{\nano\Molar}, and converting them to $pIC_{50}$ values. If there are several measurements for a kinase-compound pair, then the most active value, i.e., the entry with the highest $pIC_{50}$ value, is kept. Finally, the $pIC_{50}$ values are binarized using a $6.3$ cutoff to discriminate between an active or inactive compound as described in~\cite{Merget_2017_JMedChem}.

In the pipeline, one can additionally compute the non-reduced ratio of number of active compounds against the total number of compounds to gain insight into the actual number of measurements for each kinase pair.

@@

\subsubsection{Heatmaps}
The heatmaps are generated using matplotlib~\cite{Hunter_2007_IEEE} to depict the similarity between a set of kinases. The maximum value is $1$, indicating exact similarity, as is the case for diagonal entries. The value $0$ indicates total dissimilarity. Plotting such figures allows to see and extract patterns thanks to the gradient of colors, see top row in Figure~\ref{fig:comparison}.

\subsubsection{Dendrograms}
Clustering algorithms are used to identify groups such that the similarities within clusters are higher than compared to other clusters~\cite{Hastie_2009_ESLii}. In this study, hierarchical clustering is used, and, unlike heatmaps, it is based on distance (or dissimilarity). Hierarchical clustering can be graphically displayed using a dendrogram (see bottom row in Figure~\ref{fig:comparison}), where the height of each node is proportional to the dissimilarity between its two daughter clusters. The clustering and plotting is done using scikit-learn~\cite{Pedregosa_2011_JMLR} and matplotlib~\cite{Hunter_2007_IEEE}, respectively.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{all_comparison.png}
    \caption{Visualization of kinase similarity from four different angles: sequence, KiSSim, interaction fingerprint~(ifp) as well as ligand-profile. The top, bottom row shows four heatmaps, dendrograms respectively for a set of nine study kinases.}
    \label{fig:comparison}
\end{figure*}

@@

In the pipeline presented herein, we implemented the different methods once and streamlined each method's results into a standardized output with a pre-defined set of visualization tools for easy comparison. Moreover, the pipeline is flexible enough so that adding new methods or new visualization tools is effortless, making the whole process easy to understand, maintain, and expand.

\subsection{Means of the pipeline}
The proposed pipeline is a collection of six Jupyter notebooks~\cite{Kluyver_2016_Jupyter} that allows the study of kinase similarity from four different angles in an automated and modular fashion (Figure~\ref{fig:pipeline}).

@@

\subsection{Structure of the notebooks}
The structure of all notebooks is as follows: the first section covers the theory written in Markdown and summarizes the necessary concepts to understand the task. Relevant references are also mentioned. The second part of a notebook deals with the actual implementation of the task in a pedagogical manner, including motivation for practical steps and detailed comments on coding decisions. Finally, a discussion and a quiz section wrap up the notebook.
This structure is very well suited from a teaching perspective, since it contains both theory and hands on programming. The notebook can easily be used as a medium for a presentation, and it allows for self-study as well as usage in own research projects.

\subsection{About the code}
% Python and pythonic code
The programming section is done in Python exclusively and the code follows the latest software best practices. It is written pythonically and contains lots of code comments.
% reproducibility
Thanks to the continuous integration~(CI), all outputs and results are fully reproducible and the maintenance of the pipeline is facilitated.

@@

The final notebook collects the information from the previous ones and compares the different perspectives with easy-to-understand visualization such as heatmaps and dendrograms (see Section~\ref{sec:comparison}). Additionally, an equally weighted average to combine distance and similarity matrices from all four perspectives can be computed, yielding a single heatmap, and a single dendrogram. The user can easily extend this to a knowledge-informed weighting scheme based on their own research focus.
